A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-31 paper.tex
Tweaks to intro and refs

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output (see the
``\nameref{ARG_inference_methods}'' section), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

% \section*{Ancestral graphs and stochastic processes}
% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
% hudson1983testing, tajima1983evolutionary}
% provides the theoretical underpinning for much of contemporary population genetics.
% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
% recombination
% and the model has proved to be an especially powerful basis for
% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
% \cite{kelleher2016efficient} provide a description of his algorithm along with
% an efficient implementation, the details of which turn out to be highly relevant to
% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

% Although Hudson's algorithm was accompanied by a graph
% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
% we will use ARG to refer to a graph structure, and refer to the
% process as the Griffiths process or the coalescent with recombination
% \citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
% In the Griffiths formulation, each edge in the graph corresponds to an extant
% lineage and nodes are events in the process (the initial $n$ leaf nodes are
% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
% into a common ancestor lineage. Recombination events happen at a rate of
% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
% node, record the breakpoint and start two new edges from this node. The process
% then continues until there is only one lineage left (the Grand Most Recent
% Common Ancestor, GMRCA), which is guaranteed to
% happen in finite time because of the linear vs quadratic rates of branching
% and coalescing. Later, we will show how the graph structure and breakpoints
% associated with
% recombination nodes provides sufficient information to later recover the marginal
% trees and calculate sampling probabilities. However, this process traces out
% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
% and is consequently considerably less efficient \citep{wiuf1999ancestry}.

% Similar graph encodings have also been introduced to model other mechanisms
% which change the local ancestral tree along the genome, such as gene
% conversion~\citep{wiuf2000coalescent} and
% horizontal gene transfer~\citep{baumdicker2014infinitely}.
% [Mention the
% SMC~\citep{mcvean2005approximating,marjoram2006fast}
% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

@@

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output (see the
``\nameref{ARG_inference_methods}'' section), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{Ancestral recombination graphs}

The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
recombination
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
\cite{kelleher2016efficient} provide a description of his algorithm along with
an efficient implementation, the details of which turn out to be highly relevant to
later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

Although Hudson's algorithm was accompanied by a graph
schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
we will use ARG to refer to a graph structure, and refer to the
process as the Griffiths process or the coalescent with recombination
\citep{minichiello2006mapping}.

NOTE: it might be helpful to have a drawing of the ARG process
in-progress here to help us talk about it. What is the full state
that is remembered as we go through the process? It does help to think
about the process as choosing edges to merge, not nodes.
We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

Similar graph encodings have also been introduced to model other mechanisms
which change the local ancestral tree along the genome, such as gene
conversion~\citep{wiuf2000coalescent} and
horizontal gene transfer~\citep{baumdicker2014infinitely}.
[Mention the
SMC~\citep{mcvean2005approximating,marjoram2006fast}
and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

@@

TODO mention \citep{schaefer2021ancestral} somewhere.
